---
title: "Incubation Phase"
description: "Parmetric survival model for incubation phase durataion"
format:
  html: 
    df-print: kable
    code-fold: show
    code-summary: "Hide code"
    code-overflow: wrap
    toc-title: Page Contents
    toc: true
    toc-depth: 2
    toc-location: right
    number-sections: false
    html-math-method: katex
    smooth-scroll: true
editor: source
editor_options: 
  chunk_output_type: console
---
```{=html}
<style type="text/css">

body, td {
   font-size: 13pt;
}
code.r{
  font-size: 9pt;
}
pre {
  font-size: 11pt
}
</style>
```

```{r, warning=FALSE, message=FALSE, echo=FALSE}
library(tidyverse)
options(dplyr.summarise.inform = FALSE)

library(INLA)
library(here)
library(pals)

## Custom Functions
source(here("R/utilities.R"))
source_dir(here("R"))

## Read Data  
antem_df <- read_csv(here("local/bov_antemortem_2024.csv"))

# minimum date
min_date <- min(antem_df$date)

# add variables 
antem_df <- antem_df %>%
  mutate(hpdi = as.numeric(difftime(date, min_date, units = "hours")), # hours post donor inoculation 
         hpe = dpe*24, # convert dpe to hpe
         exp_type = if_else(group == "donor", "inoc", "cont"), # exposed by inoculation or direct contact
         fever = if_else(temp >= 104, "fever", "no_fever"), # temp >= 104 constitutes fever
         censor_status = if_else(group == "Group 1" | animal == "BR23-24", 0, 1), # No obs symptoms (0) in these 
  )

```

## Clinical Onset
The *find_clinical_onset()* function identifies first occurrence of score > 0 then creates new "Event" column with 1 at this date, 0's before this date, and a value 3 after that date. 
```{r}
clin_start_df <- as.data.frame(
  find_clinical_onset(antem_df)
)
```

Quick stats to check data
```{r}
unique(clin_start_df$Event)

check_stats <- clin_start_df %>%
  filter(Event == 1 & censor_status == 1)

dim(check_stats)
range(check_stats$dpe)
mean(check_stats$dpe)
median(check_stats$dpe)
```
  
## Remove Group 1
Group 1 was not infected and has no expectation of developing disease.  Animals that were infected, verified by nasal, serum, or fever but never developed lesions are censored.  
```{r}
clin_start_df <- clin_start_df %>%
 filter(group != "Group 1",
         Event == 1 | censor_status == 0 & censor_k == 1)
```

check data
```{r}
clin_start_df %>%
  filter(Event == 1) %>%
  summarise(mean_dpe = mean(dpe),
            median_dpe = median(dpe))
```
  
## Scale Time
Add an arbitrarily small value to eliminate zeros.  Really not need with his specific data set, by an important step.  
```{r}
clin_start_df$scaled_duration <- clin_start_df$dpe + 0.0001
```
  
## Weighted Contacts
Contact with groups by donors was sequential at 24hr intervals, shedding rates (nasal swabs) varied over this period. 
```{r}
shed_rates <- antem_df %>%
  filter(group == "donor") %>%
  group_by(dpi) %>%
  summarise(tot_shed = sum(ifelse(nasal == 45, 0, nasal))) %>%
  mutate(group = paste("Group", dpi, sep=" "))

# match cumulative shed from donors based on time of contact
clin_start_df$donor_shed <- with(shed_rates,
                                 tot_shed[match(
                                   clin_start_df$group,
                                                 group)])

# an integer index is needed for modeling purposes, group contacts are 1:1 correlated with time
clin_start_df$shed_time <- as.integer(as.factor(clin_start_df$group))
```

## Response Variable
Creating a survival object to ensure time and censored animals are correctly indicated.
```{r}
surv_obj <- inla.surv(clin_start_df$scaled_duration, clin_start_df$Event)
```

## Survival Model
```{r}
return_quants <- c(0.025, 0.05, 0.25, 0.5, 0.75, 0.95, 0.975)

pc_prec_iid <- list(theta = list(prior="pc.prec", 
                                 param=c(0.5, 0.001)))

incubation_dur <- inla(surv_obj ~ 1 +
                         f(shed_time, donor_shed,
                           model = "rw1",
                           constr=TRUE,
                           hyper=pc_prec_iid),
                        data = clin_start_df, 
                        verbose=FALSE,
                        quantiles = return_quants,
                        family = "exponential.surv",
                        control.fixed = list(prec = 1, prec.intercept  =0.0001),
                        control.compute=list(dic = TRUE, cpo = FALSE, waic = TRUE))

```

## Sample Marginals
Performing sampling on the model results
```{r}
incubation_samples <- compute_survival_marginals(incubation_dur, 14)
```

Check estimates at the 0.5 probability (median)  
```{r}
median_incubation <- find_closest_quant(incubation_samples, 0.5)
median_incubation
```

## Survival Curve
```{r fig.width=8, fig.height=6}
plot_survival_marginals(incubation_samples, x_max = 14, xlabel = "Incubation Phase Duration")
```

save incubation samples
```{r eval=FALSE}
saveRDS(incubation_samples, here("assets/incubation_samples.rds"))
saveRDS(median_incubation, here("assets/incubation_median.rds"))
```

