---
title: "Data Preprocessing"
description: "Preparing data collected during animal experiment and ANOVA"
format:
  html: 
    df-print: kable
    code-fold: show
    code-summary: "Hide code"
    code-overflow: wrap
    toc-title: Page Contents
    toc: true
    toc-depth: 2
    toc-location: right
    number-sections: false
    html-math-method: katex
    smooth-scroll: true
editor: source
editor_options: 
  chunk_output_type: console
---
```{=html}
<style type="text/css">

body, td {
   font-size: 13pt;
}
code.r{
  font-size: 9pt;
}
pre {
  font-size: 11pt
}
</style>
```

## Libraries  
Load needed libraries and packages.  
```{r, warning=FALSE, message=FALSE}
library(tidyverse)
library(here)
library(pals)
```

## Custom Functions 
Load customized functions.  
```{r}
source(here("R/utilities.R"))
source_dir(here("R"))
```

## Read Data  
Data collected during animal experiment.
```{r warning=FALSE, message=FALSE}
antem_df <- read_csv(here("local/bov_antemortem_2024.csv"))

# minimum date, donor inoculation
min_date <- min(antem_df$date)

# add variables 
antem_df <- antem_df %>%
  mutate(hpdi = as.numeric(difftime(date, min_date, units = "hours")), # hours post donor inoculation 
         hpe = dpe*24, # convert dpe to hpe
         exp_type = if_else(group == "donor", "inoc", "cont"), # exposed by inoculation or direct contact
         fever = if_else(temp >= 104, "fever", "no_fever"), # temp >= 104 constitutes fever
         censor_status = if_else(group == "Group 1" | animal == "BR23-24", 0, 1), # No obs symptoms (0) in these 
  )
```


## Plots to Check Data  

### Donors  
```{r fig.width=10, fig.height=10, warning=FALSE, message=FALSE}
plot_donors(antem_df)
```



### Contact Groups
```{r fig.width=10, fig.height=10, warning=FALSE, message=FALSE}
plot_contact_groups(antem_df)
```
  
## ANOVA  
To evaluate the differences in nasal, serum, and lesion score among groups and over time, repeated measures ANOVAs were applied using the *ezANOVA()* function from the **ez** R-package. The analyses included within-subjects (days post-exposure, dpe) and between-subjects (group) factors, with nasal, serum, and lesion scores as the dependent variables. Mauchly's test was conducted to test the assumption of sphericity. Where violations were found, Greenhouse-Geisser and Huynh-Feldt corrections were applied.  
  

**Prepare Data**    
```{r}
contact_groups <- antem_df %>%
  filter(group %in% c("Group 2", "Group 3", "Group 4")) %>%
  select(animal, group, nasal, serum, dpe, score) %>%
  mutate(
    score = replace_na(score, 0)) %>%
  drop_na() %>%
  mutate(
    score = replace_na(score, 0),          
    nasal = replace(nasal, nasal == 45, 0), # 45 indicates no detection
    serum = replace(serum, serum == 45, 0),
    dpe = as.factor(dpe),                   
    animal = as.factor(animal),            
    group = as.factor(group)                
  ) %>%
  as.data.frame()
```

::: panel-tabset

## Nasal  
Results indicate that nasal virus quantity differed significantly between the groups and across the days post-exposure, with an interaction effect suggesting that the temporal pattern of virus quantity varied among the groups.
```{r}
nasal_anova <- ez::ezANOVA(data = contact_groups, dv = .(nasal), wid = .(animal), within = .(dpe), between = .(group))
nasal_anova
```

## Serum  
Results suggest that while there were no significant differences between groups, there were significant changes over time, and the pattern of these changes differed significantly among the groups.
```{r}
serum_anova <- ez::ezANOVA(data = contact_groups, dv = .(serum), wid = .(animal), within = .(dpe), between = .(group))
serum_anova
```

## Score  
Results indicate that lesion scores differed significantly between the groups and across the days post-exposure, with an interaction effect suggesting that the temporal pattern of lesion severity varied among the groups. 
```{r}
score_anova <- ez::ezANOVA(data = contact_groups, dv = .(score), wid = .(animal), within = .(dpe), between = .(group))
score_anova
```

:::






